Higher-Order unification for free!

### 2024-04-15 paper/main.tex
section 7

Given a \linkbeta $l$ of the form \linkbetaM{\Gamma}{T}{\rhsBeta}, we provide
$3$ different activation rules:

\begin{definition}[\progBetaLL]
  Given a substitution $\sigma$, where $\sigma t_1$ is a name, say $t$, and
  $t\notin s_1\dots s_n$. If $m = 0$, then $l$ is removed and \lhs is
  unified with \rhsBetaHead. If $m > 0$, then $l$ is replaced by a
  refined version $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n,t}\appsep t_2\dots
  t_m}$ with reduced list of arguments and $Y$ being a fresh variable. Moreover,
  the new link \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}}
  is added to \linkStore.

@@

\begin{definition}[\progBetaRH]
  $l$ is removed from
  \linkStore if \rhsBetaHead is instantiated to a term $t$ and the
  $\beta$-reduced term $t'$ obtained from the application of $t$ to
  $l_1\dots l_m$ is in \llambda. Moreover, $X$ is unified with $t$.
  \label{def:progBetaRH}
\end{definition}

@@

\begin{definition}[\progBetaFail]
  If it exists a link $l' \in \linkStore$ with same \lhs as $l$, or the \lhs 
  of $l$ become rigid, then unification fail.
  \label{def:progBetaFail}
\end{definition}

@@

\begin{theorem}[Fidelity with \linkbeta]
  The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
\end{theorem}

\begin{proof}[Proof sketch]
  Let \hoUnifPb a unification problem and $\sigma$ a substitution
  such that $\hoUnifPb \in \notllambda$. 
  If $\sigma\hoUnifPb$ is in \llambda, then by
  \cref{def:progBetaLL,def:progBetaRH}, the \linkbeta associated to the subterm of
  \hoUnifPb have been solved and removed. The
  unification is done between terms in \llambda and by \cref{lemma:fidelity-maybeeta}
  fidelity is guaranteed. If $\sigma\hoUnifPb$ is in \notllambda,
  then, by \cref{def:progBetaFail}, the unification fails, as per the
  corresponding unification in \Fo{}.
\end{proof}

@@

\noindent
Initially the \linkbeta \rhs is a variable $D$ applied to the $E_x$.
The first unification problem results in $\sigma =
\{A \mapsto \lambda x.x\}$. In turn this instantiation
triggers $\linkStore_1$ by \progressetaleft 
and $E_x$ is assigned to $x$.
Under this substitution the \linkbeta becomes
\linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
it is replaced with the link:
\linketaM{}{E}{\lambda x.D_x}, while $C_x$ is unified with $D_x$. The second unification
problem assigns $f$ to $B$, that in turn activates
the second \linketa ($f$ is assigned to $C$), and then all the remaining links
are solved. The final \Ho substitution is $\sigma = \{A _{} \mapsto \lambda x.x, 
B _{} \mapsto f, C _{x} \mapsto (f\appsep x), 
D _{} \mapsto f, E _{x} \mapsto x, F _{x} \mapsto C_{x} \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
Y \mapsto f\}$.

\paragraph{Example of \progBetaRH}
We can take the example provided in \cref{sec:beta}. The problem is compiled
into:
%
% make test ONLY=7003 TEX=tex
\printAlll
  {{{X,\lambda x.Y},
    {(X\appsep a),a}}}
  {{{A,\lambda x.B},
    {C,a}}}
  {{{Y,B,0},
    {X,A,0}}}
  {{{\llambda,,C,(A\appsep a)}}}

@@

The first unification problems is solved by the substitution $\sigma = \{A
\mapsto \lambda x.B\}$. The \linkbeta becomes
\linkbetaM{}{C}{((\lambda x.B) \appsep a)} whose \rhs can be $\beta$-reduced to
$B$. $B$ is in \llambda and is unified with $C$. The resolution of the second
unification problem
gives the final substitution $\sigma = \{A \mapsto \lambda x.B, B \mapsto C, C
\mapsto a\}$ which is decompiled into $\rho = \{X \mapsto \lambda x.a, Y \mapsto
a\}$.

\subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}

\newcommand{\progBetaNoLLWait}{\emph{progress-beta-\notllambda}}

% from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
Working with terms in \llambda is sometime too restrictive \cite{Abel2018ExtensionsTM}
and a there exist a few strategies to go bejond \llambda without 
implementing Huet's algorithm~\ref{REF}.

Some implementations
of $\lambda$Prolog~\cite{lamProlog} 
such as Teyjus~\cite{Nadathur2001} delay the resolution of \notllambda unification problems until the substitution
 put them in \llambda. Other systems,
for example of the unification algorithm of Coq
used in its type-class solver~\cite{sozeau08},
 apply heuristics like preferring projection over mimic, and commit
 to that solution.

In this section we show how we can implement these strategies
by simply adding (or removing) rules to the
\elpiIn{progress} predicate.
In the example below $\foUnifPb_1$ is in \notllambda.
%
% make test ONLY=7005 TEX=tex
\printAlll
  {{{(X\appsep a),a},
    {X,\lambda x.a}}}
  {{}}
  {{}}
  {{}}

If we want the object language unification to delay the first unification problem
(waiting for $X$ to be be instantiated by a future unification),
we can relax \cref{def:progBetaFail}. 
Instead of failing because the \lhs of the considered \linkbeta $l$ becomes rigid,
we keep it in \linkStore until the head of its \rhs also become rigid.
In this case, since \lhs and \rhs have rigid heads, they can be unified
just before removing $l$ from \linkStore. We can note that this rule trivially
guarantees \fullref{prop:fidelity}. On the other hand, the occur check becomes 
partial since we break \fullref{inv:linklhs}. %: there exists \linkbeta with a non-flexible \lhs.

If we want to follow the second strategy and pick an
arbitrary solution. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
argument of the two terms is the same and unification can succeed by assigning
$X a$ to $Y$. Note
that this is not the only solution, since $\sigma = \{X \mapsto \lambda x.Y \}$
is another valid substitution. Our unification procedure
can be modified to accommodate this behavior by adding a
rule to \elpiIn{progress} that tries to
align the rightmost arguments and unify
the resulting heads whenever \lhs of a \linkbeta is not in \llambda.

Note that delaying unification outside \llambda can leave \linkbeta for the
decompilation phase. Therefore \elpiIn{commit-links} should
be modified accordingly.

% \noteD{Paragraph for decompile in case of \maybebeta}
% \todo{cita paper su estensioni a llam via sospensione etc}
\todo{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}

>>>>

Given a \linkbeta $l$ of the form \linkbetaM{\Gamma}{T}{\rhsBeta}, we provide
the following activation rules:

\begin{definition}[\progBetaLL]
  Let $\sigma$ be a substitution such that $\sigma t_1$ is a name $s$
  not occurring$s_1\dots s_n$. If $m = 1$, then $l$ is removed and \lhs is
  unified with $X_{s_1\ldots s_n~s}$.
  \todo{I changed this part, it is OK?}
  If $m > 1$, then $l$ is replaced by the link
  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n~s}\appsep t_2\dots t_m}$
  (where $Y$ is a fresh variable of arity $n+1$) and link
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n~x}}
  is added to \linkStore.

@@

\begin{definition}[\progBetaRH]
  Link $l$ is removed from
  \linkStore if \rhsBetaHead is instantiated to a term $t$ and
  $t \appsep t_1 \ldots t_m$ $\beta$-reduces to a $t' \in \llambda$.
  \label{def:progBetaRH}
\end{definition}

@@

\begin{definition}[\progBetaFail]
  \progress fails whenever:
  there exists another link $l' \in \linkStore$ with same \lhs as $l$;
  or the \lhs  of $l$ become rigid.
  \label{def:progBetaFail}
\end{definition}

\noindent In order to accommodate for the implementation of heuristics we
shall relax these failure conditions in~\cref{sec:heuristics}.

@@

\begin{theorem}[Fidelity with \linkbeta]
  The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
  if \Uo{} fails outside \llambda.
\end{theorem}

\begin{proof}[Proof sketch]
  Let $\foUnifPb_i$ be the first problem in \foUnifPb not in \llambda and
  le $\sigma$ be the substitution obtained solving $\hoUnifPb_1\ldots\hoUnifPb_{i-1}$.
  If $\sigma\hoUnifPb_i$ is in \llambda, then each variable that the
  compiler used to replace a problemat term is assigned and hence
  by \cref{def:progBetaLL,def:progBetaRH} the associated \linkbeta
  been solved and removed. Note that in this case all
  calls to \Ue{} are between terms in \llambda and by \cref{lemma:fidelity-maybeeta}
  fidelity is guaranteed. If $\sigma\hoUnifPb$ is still in \notllambda,
  then by \cref{def:progBetaFail} unification fails as the
  corresponding unification in \Fo{} would.
\end{proof}

@@

\noindent
Initially the \linkbeta \rhs is a variable $D$ applied to the $E_x$.
The first unification problem results in $\sigma =
\{A \mapsto \lambda x.x\}$. In turn this instantiation
triggers $\linkStore_1$ by \progressetaleft 
and $E_x$ is assigned to $x$.
Under this substitution the \linkbeta becomes
\linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
it is replaced with the link:
\linketaM{}{E}{\lambda x.D_x}, while $C_x$ is unified with $D_x$. The second unification
problem assigns $f$ to $B$, that in turn activates
the second \linketa ($f$ is assigned to $C$), and then all the remaining links
are solved. The final \Ho substitution is $\sigma = \{~A _{} \mapsto \lambda x.x ~;~
B _{} \mapsto f ~;~ C _{x} \mapsto (f\appsep x) ~;~
D _{} \mapsto f ~;~ E _{x} \mapsto x ~;~
F _{x} \mapsto C_{x} ~\}$ and is decompiled into $\rho = \{~X \mapsto \lambda x.x ~;~
Y \mapsto f~\}$.

\paragraph{Example of \progBetaRH}
We give a second look at the opening example in \cref{sec:beta}.
\todo{this is a little more complex that the opening problem}
The problem is compiled into:
%
% make test ONLY=7003 TEX=tex
\printAlll
  {{{X,\lambda x.Y},
    {(X\appsep a),a}}}
  {{{A,\lambda x.B},
    {C,a}}}
  {{{Y,B,0},
    {X,A,0}}}
  {{{\llambda,,C,(A\appsep a)}}}

@@

The first unification problems is solved by the substitution
$\sigma = \{~A \mapsto \lambda x.B~\}$. The \linkbeta becomes
\linkbetaM{}{C}{((\lambda x.B) \appsep a)} whose \rhs can be $\beta$-reduced to
$B$. $B$ is in \llambda and is unified with $C$. The resolution of the second
unification problem
gives the final substitution $\sigma = \{~A \mapsto \lambda x.B ~;~ B \mapsto C ~;~ C
\mapsto a ~\}$ that is decompiled into $\rho = \{~ X \mapsto \lambda x.a ~;~ Y \mapsto
a ~\}$.

\subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}
\label{sec:heuristics}
\newcommand{\progBetaNoLLWait}{\emph{progress-beta-\notllambda}}

% from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
Working with terms in \llambda is sometime too restrictive~\cite{Abel2018ExtensionsTM}
and we could find in literature a few strategies to go beyond \llambda without 
implementing Huet's algorithm~\cite{Huet75}.
Some implementations of $\lambda$Prolog~\cite{lamProlog} such as
Teyjus~\cite{Nadathur2001} delay the resolution of \notllambda unification
problems until the substitution makes them reenter \llambda. Other systems, for
example of the unification algorithm of Coq used in its type-class
solver~\cite{sozeau08}, apply heuristics like preferring projection over mimic,
and commit to that solution.

In this section we show how we can implement these strategies
by simply adding (or removing) rules to the
\elpiIn{progress} predicate.
In the example below $\foUnifPb_1$ is in \notllambda.
%
% make test ONLY=7005 TEX=tex
\printAlll
  {{{(X\appsep a),a},
    {X,\lambda x.Y}}}
  {{{A,a},
    {B,\lambda x.C}}}
  {{{Y,C,0},
    {X,B,0}}}
  {{{\llambda,,A,(B\appsep a)}}}

\noindent
If we want the object language unification to delay the first unification problem
(waiting for $X$ to be be instantiated),
we can relax \cref{def:progBetaFail}. 
Instead of failing when the \lhs of $\linkStore_1$ becomes rigid (equal to $a$),
we keep it in \linkStore until the head of its \rhs also become rigid.
In this case, since both the \lhs and \rhs have rigid heads, they can be unified.
While this relaxed rule does not break \fullref{prop:fidelity} per se,
the \elpiIn{occur-check-links} procedure becomes incomplete since
\fullref{inv:linklhs} is broken. Also note that delaying unification outside \llambda can leave \linkbeta for the
decompilation phase. Therefore \elpiIn{commit-links} should
be modified accordingly.

If instead we want \Uo{} to follow the second strategy and pick an
arbitrary solution we can modify \progress by applying the desired heuristic.
For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
argument of the two terms is the same and unification can succeed by assigning
$X a$ to $Y$. This heuristic can be implemented by adding the following rule:

\input{code/progress_heuristic}

\noindent
Note that \elpiIn{UV} is in \llambda as well as \elpiIn{T}.
